---
title: "Asset Pricing - PS-1"
author: "Guneykan Ozkaya, Dominic Henderson, Mardel Faria"
date: "October 05, 2019"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


\def\Cov{\mathrm{Cov}}

Question 1

1-a)For each $\gamma$ compute the value of $m$ in each year. Use then the entire time series of $m$
and $R_M$ to compute  $-\Cov(R_M, m)$ and multiply it by the mean of $1+r$. Plot the
resulting values against all values of $\gamma$ Which of those values comes closest to the
model’s theoretical prediction (mean value of $R_M-r$)?

For each $\gamma$,
\[
m = \frac{U'(c_1)}{U'(c_0)} = \left( \frac{c_0}{c_1} \right)^\gamma
\]
and
\[
E(R_m-R_f) = -\Cov(m, R_m) (1 + R_f) .
\]

```{r}

library("readxl")
library("data.table")
library("tinytex")
#df <- read_excel("/Users/guneykan/Desktop/PS1Data.xlsx")
df <- read_excel("/Users/guneykan/Desktop/PS1Data.xlsx")
df<-data.table(df)
names(df)<-c("year", "dc", "r_p", "r_f")
```

```{r}
hist_Rp<-df[, mean(r_p)]
hist_Rf<-df[, mean(r_f)]
r_m<-df[, r_p] + df[, r_f] - 1
df[, r_m:=list(r_m)]
df[, m:=list(1/df[, dc])]
gamma<-c(2:100)
m_gamma<-vector("list", 99)
cov_r_m<-vector("list", 99)
c_capm_Rp<-vector("list", 99)
for(i in c(1:99)){
  m_gamma[[i]]<-(df[, m])^gamma[i]
  cov_r_m[[i]]<-cov(m_gamma[[i]], df[, r_m])
  c_capm_Rp[[i]]<-cov_r_m[[i]]*hist_Rf*-1
}


```

```{r}
plot(gamma, unlist(c_capm_Rp),
main="C_CAPM Rp",
ylab="Rp",
xlab = "Gamma",
type="l",
col="blue")
lines(gamma, rep(hist_Rp, 99), col="red")
index<-which.min(abs(unlist(c_capm_Rp)-hist_Rp))
cat("closest CAPM estimation:", unlist(c_capm_Rp)[index]," ",
    "mean Rp:", hist_Rp," ",
    "respective gamma:", gamma[index]) 

```



1-b) For each $\gamma$ compute the inverse of the sample mean of $m$ and plot against all values
of $\gamma$ Which of those values comes closest to the model’s theoretical prediction
(mean value of $1 + r$))?

The last part means that we need to find a value of $\gamma$ such that
\[
1 + R_f = 1/E(m) .
\]


```{r}

mu_m<-sapply(m_gamma, mean)
c_capm_rf<-1/mu_m 
plot(gamma, c_capm_rf,
main="C_CAPM Rf",
ylab="Rf",
xlab = "Gamma",
type="l",
col="blue")
lines(gamma, rep(hist_Rf, 99), col="red")
index_0<-which.min(abs(c_capm_rf-hist_Rf))
cat("closest CAPM estimation:", c_capm_rf[index_0], " ", 
    "mean Rf:", hist_Rf, " ", 
    "respective gamma:", gamma[index_0]) 

```


1-c) What conclusions do you draw from these results?

Replacing $m$ with the Taylor Expansion $m \approx 1 - \gamma \frac{c_1 - c_0}{c_0}$ we obtain
\[
E(R_m-R_f) = \gamma \Cov\left( \frac{(c_1-c_0)}{c_0}, R_m \right) (R_f+1)
\]

With these preferences the Relative Risk Aversion Coefficient is constant,
\[
R(c) = - \frac{U''(c)}{U'(c)} c  = \frac{\gamma}{c} c = \gamma
\]
We see that, as $\gamma$ increases, investors become more risk-averse. Thus the expected risk premium increases.

The relative risk aversion coefficient is equal to $\gamma$ for all investors, so it makes sense to compare the C-CAPM to aggregate data. Comparing the data with the implications of the Consumption CAPM model, $\gamma$ needs to be around 87 for the C-CAPM model to match the historical risk premium, meaning investors need to be extremely risk-averse. This is the equity premium puzzle.

On the other hand, when we look at the estimation of the risk-free rate by the C-CAPM model, $\gamma$ needs to be around 2 to match the historical risk-free rate. These two estimations of the model do not match with each other, or the data itself. While the estimated risk premium indicates that investors need to be highly risk-averse to explain the empirical risk premium, comparison of the historical risk-free and the model estimation shows $\gamma$ needs to be around 2, which is the complete opposite of the result we obtained at 1(a).

We see that as investors become more risk-averse the covariance of the stochastic discount factor $m$, and the market return $R_M$ goes more negative. This is because the covariance depends on the risk aversion of the inverstors and the expectation of the future consumption of investors(c_1). As investors become more risk-averse or as their expectation of the future consumption decrease(probability of bad state increase), they will require a higher risk premium to compensate for the risk they are taking by holding the portfolio.

Also, the model estimation of risk-free shows that as the investors become more risk-averse ($\gamma$ increases), the risk-free rate also increases. This may not make sense if one thinks that if the investors are more risk-averse, they tend to invest more in bonds. This will push the price of bonds up and bringing the risk-free rate down. What we see here is the opposite. But this can be seen from the different viewpoint. Because we assume investors are not impatient ($\delta = 1$) and because in addition to risk awareness, $\gamma$ also reflects how much investors dislike consumption fluctuations (inverse of intertemporal substitution elasticity), as $\gamma$ goes up, risk premium goes up meaning investors are compensated for the risk they are taking and the market is in equilibrium (market compensates investors against the fluctuations in consumption growth) thus demand for the bonds decrease, and the risk-free rate increases (risk-free assets).





Question 2

There is no question 2(a)

2-b) Stochastic discount factor

\[
U(c_t) = \left( \frac{c_t}{x_t-1} \right)^{1-\gamma}
\quad\implies\quad
U'(c_t) = \frac{c_t^{-\gamma}}{x_{t-1}^{1-\gamma}}
\quad\text{and}\quad
U'(c_{t-1}) = \frac{c_{t-1}^{-\gamma}}{x_{t-2}^{1-\gamma}}
\]

Thus the SDF is:
\begin{align*}
m &= \frac{U'(c_t)}{U'(c_{t-1})}=\left(\frac{c_t}{c_{t-1}}\right)^{-\gamma}\left(\frac{x_{t-1}}{x_{t-2}}\right)^{\gamma-1}
\\[4pt]
  &= \left( \frac{c_t}{c_{t-1}} \right)^{-\gamma}\left(\frac{c_{t-1}}{c_{t-2}}\right)^{\gamma-1}
\end{align*}

```{r}
m_gamma_abel<-vector("list", 99)
cov_r_m_abel<-vector("list", 99)
c_capm_Rp_abel<-vector("list", 99)
abel_m<-c(NA, df[2:63, m])
df[, abel_factor:=list(abel_m)]
for(t in c(1:99)){
  for(i in c(2:63)){
    m_gamma_abel[[t]][i-1]<-((df[(i-1), m])^(1-gamma[t]))*(df[i, abel_factor]^(gamma[t]))
  }
}
for(i in c(1:99)){
  cov_r_m_abel[[i]]<-cov(m_gamma_abel[[i]], df[2:63, r_m])
  c_capm_Rp_abel[[i]]<-cov_r_m_abel[[i]]*hist_Rf*-1
}


```


2-c) Risk Premium estimation for each gamma


```{r}
plot(gamma, unlist(c_capm_Rp_abel),
main="C_CAPM Rp, Abel",
ylab="Rp",
xlab = "Gamma",
type="l",
col="blue")
lines(gamma, rep(hist_Rp, 99), col="red")
index<-which.min(abs(unlist(c_capm_Rp_abel)-hist_Rp))
cat("closest CAPM estimation:", unlist(c_capm_Rp_abel)[index]," ",
    "mean Rp:", hist_Rp," ",
    "respective gamma:", gamma[index]) 
```


2-d) Risk-Free rate estimation for each gamma


```{r}
mu_m_abel<-sapply(m_gamma_abel, mean)
c_capm_rf_abel<-1/mu_m_abel
plot(gamma, c_capm_rf_abel,
main="C_CAPM Rf, Abel",
ylab="Rf",
xlab = "Gamma",
type="l",
col="blue")
lines(gamma,rep(hist_Rf, 99), col="red")
index_0<-which.min(abs(c_capm_rf_abel-hist_Rf))
cat("closest CAPM estimation:", c_capm_rf_abel[index_0], " ", 
    "mean Rf:", hist_Rf, " ", 
    "respective gamma:", gamma[index_0]) 
```


2-e) How  does  the  model  compare  relative  to  the  model  with  simpler  preferences  of  question 1? 

Considering updated utility function we obtain the relative risk aversion coefficient
\[
R(C) = -\frac{U''(c)}{U'(c)} c = \gamma
\]
This is still constant so we can still use aggregate consumption as a representative agent to see how well the model matches it.

Compared to question 1(a), now the $\gamma$ which matches the historical risk premium is around 36. While this value implies less risk aversion relative to the estimation in question 1(a), it is still a high degree of risk aversion. Also, it is still not close to the $\gamma$ we require to match the model's risk-free rate to the historical risk-free rate. Due to the fact that the utility function takes into account previous consumption growth in addition to the current consumption growth, it is possibly more realistic compared to C-CAPM, if we think actual investors also want to keep track with others' consumption, or their actual previous consumption. In the asset pricing model of Abel (1990), we see that as $\gamma$ increases (that is, as investors become more risk-averse), the risk-free rate decreases which is due to stochastic discount factor taking previous consumption growth into account in addition to current consumption growth.
